% although substantial heat saving measures are installed
\section{Introduction}
The Paris Climate Agreement sets the global framework for mitigating climate change \cite{agreement2015paris}. It stipulates that the increase in the global average temperature should be kept well below \SI{2}{\degreeCelsius} compared to 1990. In addition, further measures are developed, aiming at a maximum increase of \SI{1.5}{\degreeCelsius}. However, it is also about humanity adapting to the negative effects of climate change that are already being felt. The IPCC Special Report on \SI{1.5}{\degreeCelsius} (SR1.5) summarizes the state of scientific knowledge globally on the consequences of global warming \cite{edenhofer2011ipcc}.

\subsection{Long-term global sustainable transformation plans of energy systems}
To implement the Paris Climate Agreement and the SR1.5, the European Commission has set deep decarbonization targets together with national governments. In particular, the "EU Green Deal" describes the concrete goals in Europe, namely a climate-neutral and resource-conserving economy and society (see, e.g., in \cite{kemfert2019green}). The overarching goal is emissions neutrality 2050. To achieve this long-term ambition, the European Commission recently presented "Fit for 55", a concrete roadmap to 2030. This program commits to a \SI{55}{\%} reduction in CO\textsubscript{2} emissions in 2030 compared to 1990 \cite{european_commission_european_2019}. The concrete measures affect almost all sectors of the energy system and should lead to a significant efficiency improvement and a massive overall reduction in fossil fuels. It implies, among others, binding annual targets for reducing energy consumption and an extension of the already established EU emissions trading system (EU ETS) to new sectors. In addition to transportation, the building sector will also be part of the EU ETS in the future. A separate new emissions trading system for fuel supply in these sectors will be introduced. In the buildings sector, through the annual anchored emissions reduction, this means a defined roadmap to complete decarbonization of heating and cooling demand, as the two reasons for emissions in this sector. In this paper, we look at what deep decarbonization of building heating demand may look like in 2050 in Austria and the implications of the corresponding sustainable energy mix for centralized heating networks.

\subsection{Implications and effects of the decarbonization on the heating sector}
The scope of changes required by 2030/2050 in the heating sector become even clearer at the national level. In Europe, the average share of renewable energies in the heating and cooling sector 2018 is only just above \SI{20}{\%} on average for all EU member states \cite{eurostat_reference}. It is in fact higher in some countries, for example in Austria, where it is above \SI{34}{\%}. However, fossil fuels continue to dominate there as well. To be even more specific for the heating sector: of the nearly 4,000,000 residential dwellings in Austria, more than 900,000 are heated with natural gas, and more than 500,000 with oil \cite{statistik_austria}. If these heating systems are converted to renewable energy supply by 2050, this corresponds to a retrofitting of 50,000 units per year, or more than 130 per day - only in Austria. To achieve this goal makes measures necessary that go beyond the electrification of heat and leads to an expansion of district heating networks \cite{jalil2018spatially}.\vspace{0.3cm}

Centralized heating networks are particularly advantageous for supplying densely populated or urban areas resulting from high heat densities there \cite{inage2020development}. In addition to heat density, the connection rate is a key factor determining the efficiency of district heating/cooling networks and thus their implementation. For example, currently in Austria, at a connection rate of \SI{90}{\%}, \SI{10}{GWh \per km^2} is used as a benchmark for supplying an area with district heating\footnote{\url{http://www.austrian-heatmap.gv.at/ergebnisse/}}. The reference value of \SI{10}{GWh \per km^2} is in line with findings regarding district heating networks also from the Scandinavian region (Denmark, Sweden, and Finland) \cite{zinko2008district}. These are rough estimates, but they do allow an initial assessment of the economic viability or feasibility of a district heating network. In a detailed consideration and evaluation of district heating networks numerous factors play a decisive role. Nussbaumer and Thalmann \cite{nussbaumer2016influence} thoroughly elaborate on the network design and its impact on the profitability of centralized heat networks. Laasasenaho et al. \cite{laasasenaho2019gis} emphasize in their study the optimal location of heat generation units/sources within centralized heat networks enabling a cost-optimized heat supply. Gopalakrishnan and Kosanovic \cite{gopalakrishnan2014economic} focus in their study on the optimal heat generation technology dispatch. When examining the economic viability of district heating networks, building renovation measures must also be taken into account (see, e.g., in \cite{andric2018impact} and \cite{rabani2021achieving}). Hietaharju et al. \cite{hietaharju2021stochastic} recently show in their analysis that a $2-3$\SI{}{\%} building renovation rate per year results in a decrease of $19-28$\SI{}{\%} of the long-term district heating demand. This reduces also the heat density. However, studies show that a reduction in heat density is not necessarily a barrier to district heating networks \cite{persson2011heat}. Reidhav and Werner \cite{reidhav2008profitability} show in their study how energy taxes can improve the profitability of sparse district heating networks in Sweden. Following these considerations and in light of ambitious CO\textsubscript{2} reduction targets, it can also be assumed that the rising CO\textsubscript{2} price can have a similar effect as the energy tax. Of course, this is only valid in case of deep decarbonization of the generation mix feeding into centralized heat networks. Di Lucia and Ericsson \cite{di2014low} show that biomass significantly contributed to the decarbonization of the district heating network and replaced fossil fuels in the feed-in generation mix in Sweden. Ghafghazi et al. \cite{ghafghazi2010multicriteria} also identify in their mutli-criteria study wood pellets as the optimal system options for fueling district heating networks. Eventually, also the increasing cooling demand and the co-design of centralized networks for heating and cooling can increase the economic viability of these and counteract the reduction of heat density from an economic point of view \cite{zhang2021economic}.

\subsection{Lack in the implementation of decarbonization in different sectors}
However, the concrete implementation to achieve predefined climate change mitigation goals still is lacking in many cases. For this reason, numerous studies go beyond and show roadmaps for the rapid decarbonization. For example, Rockstr{\"o}m et al. \cite{rockstrom2017roadmap} conduct such a study and propose pathways for halving gross anthropogenic CO\textsubscript{2} emissions every decade. Other works go into more depth regarding optimal solutions for the decarbonization of individual energy services. There are relevant differences between the individual sectors of the energy system related to decarbonization. How a sustainable energy service can be provided in the different sectors must therefore be examined in detail. This perspective is supported by a large number of detailed decarbonization studies covering specific energy service needs (e.g., for the building sector Leibowicz et al. \cite{leibowicz2018optimal}, transport sector Pan et al. \cite{pan2018decarbonization}, and industries Habert et al. \cite{habert2020environmental}).\vspace{0.3cm}

Despite all the details associated with the sector-specific decarbonization strategies, the principles of a net-zero society base on three key points: (i) reduction of the energy demand (see, e.g., Oshiro et al. \cite{oshiro2021enabling} and Grubler et al. \cite{grubler2018low}), (ii) deployment and generation of renewable energy technologies (see, e.g., Bakhtavar et al. \cite{bakhtavar2020assessment} focusing on net-zero districts by deployment of renewable energy generation), and (iii) increase in efficiency regarding the provision of energy services and the associated optimal utilization of sustainable energy sources. The third point (iii) includes, among others, two main aspects, namely: on the one hand, that potentials of renewable resources are exploited locally and, on the other hand, that energy carriers with various fields of application are utilized with the highest possible efficiency. We like to refer to just a few selected references without claiming to be exhaustive and focus here on hydrogen as one example of an energy carrier with high potentials in sustainable energy systems and a significant bandwidth of efficiency in terms of its generation, storage and use. Van Ruijven et al. \cite{van2007potential} highlight that the introduction of hydrogen in global energy systems only leads to lower emissions with high end-use efficiency and low-carbon production. Van Ressen \cite{van2020hydrogen} systematically investigates the possibilities and challenges of hydrogen and discusses extensively its role in the energy transition. Recently, Böhm et al. \cite{bohm2021power} comprehensively elaborate on hydrogen-related synergies and its role in sustainable heat supply.\vspace{0.3cm}

\subsection{Implications of large-scale numerical model results on the local level}
In many cases when it comes to the question of optimal solutions, researcher use numerical models. In general, these models strike a balance between complexity and aggregation. Integrated assessment models (IAMs) are large numerical models covering complex interrelations between climate, society, economics, policy, and technology. Dowlatabadi \cite{dowlatabadi1995integrated} provided 1995 a fundamental review on IAMs focusing on their role in the context of climate change. Krey et al. \cite{krey2019looking} discuss and systematically compare different IAMs. Harmsen et al. \cite{harmsen2021integrated} elaborates on the modeling behaviour of IAMs. Wilkerson et al. \cite{wilkerson2015comparison} and van Vuuren et al. \cite{van2016carbon} deal with IAMs and their role in understanding global energy decarbonization pathways. In particular, both studies examine CO\textsubscript{2} budget and price developments. Schwanitz \cite{schwanitz2013evaluating} evaluates IAMs of global climate change and discusses, among others, the appropriate level of regional (spatial) aggregation of countries in the modeling analysis. Generalizing this aspect reveals an aspect already known but essential in the context of large numerical models. It becomes necessary for modelers to set priorities regarding the level of detail, which inevitably creates trade-offs in the analysis regarding the granularity of the temporal, spatial, and other dimensions \cite{gargiulo2013long}. Gambhir et al. \cite{gambhir2019review} also highlight this aspect of aggregation bias in their critical review of IAMs. They propose, among others, that IAMs should be increasingly be supplemented with other models and analytical approaches. Not least for this reason, (large) energy models also play a significant role in the analysis of energy systems in the context of climate change. Compared to IAMs, they more strongly emphasize the level of detail in terms of techno-economic characteristics (see the review of modeling tools of energy systems in \cite{ringkjob2018review}). However, the lack of granularity remains, that these (global) energy models consider only a highly aggregated spatial resolution. To name just two selected approaches, Capros et al. \cite{capros2012model} (PRIMES) and Löffler et al. \cite{loffler2017designing} (GENeSYS-MOD) provide energy system models focusing on the European energy system with a spatial resolution on the country level. Further approaches are needed to disaggregate results obtained at the country level to finer scales, such as districts, neighborhoods, and other local levels. In this context, Backe et al. \cite{backe2021heat} provided a novel approach in the context of merging local activities/behavior in sustainable local communities into a large energy system model (bottom-up linkage). In their study, they integrated local flexibility options into the global energy system model EMPIRE, which provides in principle only country-level resolution. This and other work confirms the emerging trend of making top-down and bottom-up linkages between different spatial-temporal levels of resolution to drive decarbonization across all sectors.\vspace{0.3cm}

\subsection{Objective and contribution of this work}
Against this background, the core objective of this work is downscaling European decarbonization scenarios of the heating sector to the community/distribution grid level serving end-users in 2050. In particular, downscaling considers the highly efficient and local use of sustainable heat sources in centralized heat networks (e.g., co-firing hydrogen in cogeneration plants and large-scale waste utilization, etc.). In addition, the topography of district heating networks is of particular importance and plays a crucial role in applied downscaling. This allows estimates of realistic decarbonized district heating networks in 2050 to be obtained, which can be compared with existing networks. Thereby, the heat density of district heating networks serves as a comparative indicator and permits a rough estimation of the changes needed of centralized heating networks considering the 1.5°C climate target. An Austrian case study is conducted, downscaling the results of the heating sector in 2050 from the large numerical energy system model GENeSYS-MOD, from the country to community/distribution grid level.\vspace{0.3cm}

The method applied consists of three different scenario-independent downscaling techniques. As the first, proportional downscaling using population as proxy is used as reference (Section \ref{pop}). As the second, an sequential downscaling approach is presented, dissaggregating from the country level to the sub-region level. Thereby, population density and the infrastructure requirements of heat technologies serve as additional criterion in the downscaling (Section \ref{alg1}). And finally, an iterative downscaling algorithm is presented. The algorithm is based on graph-theory benchmarking and projects centralized heat supply on the local (community) level (Section \ref{alg2}). Section \ref{results} presents and discusses the results of this work. Section \ref{res:1} and \ref{res:2} show heat generation by source on different spatial levels. Section \ref{res:3} and \ref{res:4} present centralized heat networks on a high spatial granularity. Section \ref{res:5} synthesizes the results of centralized heat networks and compares heat densities of centralized heat networks in 2050 with today's values. Section \ref{conclusions} concludes this work and provides an outlook for future work. 