\section{Materials and methods}\label{methodology}
This section explains the methodology \replaced{of this work.}{developed in this work.} \added{First, section \ref{res:1} presents the output from the Horizon 2020 project openENTRANCE, since this is the main input for the downscaling. Therein, information about the different heat sources/generation technologies that are downscaled and used for district heating is provided. Then, the different downscaling techniques are explained.} Section \ref{pop} describes proportional spatial downscaling using population as a proxy. Building on this, section \ref{alg1} presents the sequential downscaling and section \ref{alg2} presents the iterative downscaling algorithm in detail. \added{Section \ref{sec:limitations} discusses selected limitations of the methodology.} Finally, section \ref{open} concludes this section and explains the open-source tools used in this work.

\subsection{Heat supply of the Austrian residential and commercial sector in 2050: four different decarbonization scenarios\deleted{ obtained from the Horizon 2020 project openENTRANCE}}\label{res:1}
This section presents the heat generation mix covering the Austrian residential and commercial heat demand in 2050 for four different \replaced{scenarios}{storylines}, which have been developed within the Horizon 2020 openENTRANCE project. They are named as follows: \textit{Directed Transition}, \textit{Societal Commitment}, \textit{Techno-Friendly}, and \textit{Gradual Development}. Within each of them, specific fundamental development of the energy systems is described while aiming for a sustainable transition of the provision of energy services. The first three \replaced{scenarios}{storylines} assume different approaches to limit global warming to around \SI{1.5}{\degreeCelsius} as laid out in the Paris Agreement. \added{Particularly, the results of these scenarios implicitly consider the remaining European fraction of the CO\textsubscript{2} budget of the 1.5°C climate target.} The last \replaced{scenario}{storyline} (\textit{Gradual Development}) can be interpreted as less ambitions \replaced{scenario}{storyline}, limiting global warming to around \SI{2.0}{\degreeCelsius} climate target. \added{Accordingly, the results of this scenario consider the remaining European fraction of the CO\textsubscript{2} budget of the 2.0°C climate target.} Below, the \replaced{scenarios}{storylines} are described briefly, before the quantitative results at the country level are presented. For a more detailed description of the \replaced{scenarios}{storylines}, refer to \cite{auer2020quantitative} and \cite{auer2020development}. Further information is also available on the website of the project\footnote{\url{https://openentrance.eu/}} and on GitHub\footnote{\url{https://github.com/openENTRANCE}}.\vspace{0.3cm}

The underlying concept of the four \replaced{scenarios}{storylines} is a three-dimensional space consisting of the following parameters: technology, policy, and society. Each \replaced{scenario}{storyline} describes a specific pathway to reach a decarbonized energy system taking into account a pronounced contribution of two dimensions. Regarding the third dimension, a development is assumed that leads to no significant contribution to the decarbonization of the energy system. 

\begin{itemize}
	\item \textit{Directed Transition} looks at a sustainable provision of energy services through strong policy incentives. This bundle of actions becomes necessary because neither the markets nor the society adequately pushes sustainable energy technologies.
	\item \textit{Societal Commitment} achieves deep decarbonization of the energy system by a strong societal acceptance of the sustainable energy transition and shifts in energy demand patterns. Thereby, decentralized renewable energy technologies together with policy incentives facilitate a sustainable satisfaction of energy service needs. Due to the shift in energy demand, no fundamental breakthroughs of new clean technologies are required.
	\item \textit{Techno-Friendly} describes a development of the energy system where a significant market-driven breakthrough of renewable energy technologies gives rise to the decarbonization of energy service supply. Additionally, society acceptance supports the penetration of clean energy technologies and the sustainable transition.
	\item \textit{Gradual Development} differs from the other \replaced{scenarios}{storylines}: it assumes emissions reductions that (only) stabilize the global temperature increase at \SI{2.0}{\degreeCelsius}. At the same time, a combination of each possible sustainable development initiative of the energy system is realized in this \replaced{scenario}{storyline}. Although the other three dimensions contribute to decarbonization, they do not push it sufficiently and result in a more conservative \replaced{scenario}{storyline} than the others.
\end{itemize}

Table \ref{tab:comparison} shows the heat generation by source/technology in Austria in 2050 for the four different \replaced{scenarios}{storylines}. These values were obtained during the course of the Horizon 2020 project openENTRANCE and are the modeling results calculated using the open-source model GENeSYS-MOD \cite{burandt2018genesys}. \added{In this work, the naming convention of heat sources/generation technologies from GENeSYS-MOD is essentially followed to ensure consistency between aggregated (i.e., input values) and downscaled (i.e., output values). However, we introduced the heat sources waste and geothermal that were initially not included in the list of heat sources from openENTRANCE results. We separated waste as part of biomass and geothermal from heat pump (ground) heat generation using estimates from national Austrian studies in \cite{konighofer2014potenzial} and \cite{buchele2015bewertung}.}

\definecolor{Gray}{gray}{0.95}
\begin{table}[h]
	\centering
	\resizebox{0.9\textwidth}{!}{% use resizebox with textwidth
		\renewcommand{\arraystretch}{1.1}
		\begin{tabular}{lrrrrr}
			\toprule 
			& 2020 & \multicolumn{4}{c}{2050}\\
			\cmidrule(lr){2-2}\cmidrule(lr){3-6}
			Generation by source in TWh  & - & DT & SC & TF & GD\\\hline
			Biomass & 13.00 & 3.37 & 3.37  & 3.37  & 3.37 \\
			Direct electric & 4.10 & 2.13  & 1.98 & 1.53  & 1.81 \\
			Geothermal & 0 & 2  & 2  & 2  & 2 \\
			Natural gas (fossil) & 43.67 & 0  & 0  & 0  & 0 \\
			Heat pump (air) & 11.37 & 22.73  & 15.71  & 25.96  & 9.68 \\
			Heat pump (ground) & 0 & 17.50  & 19.47  & 4.69  & 19.21 \\
			Hydrogen & 0 & 1.03  & 2.18  & 7.43  & 8.65 \\
			Oil & 0.66 & 0  & 0  & 0  & 0 \\
			Synthetic gas & 0 & 0.36  & 1.35  & 2.79  & 5.35 \\
			Waste & 1.2 & 2  & 2  & 2  & 2 \\\hline
			\cellcolor{Gray} Total & \cellcolor{Gray}74.0 &\cellcolor{Gray}51.12 & \cellcolor{Gray}48.06&\cellcolor{Gray}49.77 & \cellcolor{Gray}52.07\\
			Rel. reduction compared to 2020& - & -31\% & -35\% & -33\% & -30\%\\\hline
			District heating &  & 5.39 & 7.53 & 14.22 & 18.00\\
			\bottomrule
	\end{tabular}}
	\caption{\added{Heat generation by source in Austria in 2020 and the four different decarbonization scenarios in 2050. Source: \cite{auer2020development},\cite{konighofer2014potenzial},\cite{buchele2015bewertung}}}
	\label{tab:comparison}
\end{table}

\added{The total heat generation (and thus total heat demand) is significantly reduced when comparing the values of 2020 and 2050. The heat demand reduction varies between -30\% and -35\% and is highest in the \textit{Societal Commitment} scenario. The row district heating (last row) describes the heat generation used in district heating networks and is the sum by geothermal, hydrogen, synthetic gas and waste. This is further and detailed described in section \ref{alg1} and Table \ref{tab:require} therein.}
\deleted{According to the underlying assumptions in the storylines, the heat generation of the different sources/technologies varies significantly in some cases (e.g., hydrogen-based heat generation in \textit{Directed Transition} and \textit{Gradual Development} (7.62TWh) or heat pump (ground) generation in \textit{Techno-Friendly} and \textit{Societal Commitment} (14.78TWh)). The gray-colored column $\Sigma$ presents the total heat generation using centralized heat networks, which varies between 19.49TWh (\textit{Techno-Friendly}) and 35.23TWh (\textit{Gradual Development}).}

\subsection{Proportional spatial downscaling using population as a proxy}\label{pop}
Proportional downscaling is a well-established technique for spatial disaggregation and is often used in scientific and practical studies. Equation \ref{eq:1} shows a mathematical formulation of proportional downscaling for disaggregation of energy demand $d$ from the country to the local levels, using population $p$ as a proxy.
\begin{align}\label{eq:1}
d_{local}=\frac{p_{local}}{p_{country}} \cdot d_{country}
\end{align}
The fields of application of proportional downscaling are not limited to the modeling of energy systems but to different fields of scientific and practical studies. The reason for this is the intuitive application and that it offers possibilities for tailor-made adaptions, in particular, related to the downscaling driver and proxy\deleted{\cite{van2006downscaling}}. In this context, \replaced{the study in \cite{van2006downscaling}}{van Vuuren et al. \cite{van2006downscaling}} provide\added{s} a comprehensive analysis of different proxies for the downscaling of global environmental change, including gross domestic product, emissions and other indicators. \deleted{However, in the context of downscaling aggregated values of energy systems, one often finds proportional downscaling using population as a proxy (see, e.g., Ahn et al. \cite{ahn2019downscaled}, van Vuuren et al. \cite{van2010downscaling}, and Alam et al. \cite{alam2018downscaling})}. \added{However, downscaling aggregated values of energy system often uses proportional downscaling and population as a proxy \cite{alam2018downscaling}.} \replaced{Further information can be found in the review study in}{For further information, we refer the reader to van Vuuren's study} \cite{van2010downscaling}, providing a systematic classification of \added{different} downscaling techniques \deleted{going far beyond the simple proportional downscaling method discussed so far}. The reader can find population-based downscaling in the authors' categorization under algorithmic and proportional downscaling. In addition, \replaced{the study shows}{they showed} that novel downscaling methods have emerged in recent years as the scientific community has increasingly recognized the necessity for spatial and temporal disaggregation.

\subsection{Sequential downscaling (from the country to the sub-region level)}\label{alg1}
The sequential approach (Algorithm 1) downscales the heat generation by source from the country to the sub-region levels. Before explaining the algorithm in detail, Table \ref{tab:nuts} provides an overview of the spatial nomenclature of this work using the European nomenclature of territorial units for statistics\footnote{\url{https://ec.europa.eu/eurostat/web/nuts/background}.} (NUTS) and gives some examples of Austria. In particular, the different spatial levels of the applied downscaling are marked in gray. According to the NUTS nomenclature, Algorithm 1 downscales from the NUTS0 level to the NUTS3 level.\vspace{0.3cm}

\definecolor{Gray}{gray}{0.95}
\begin{sidewaystable}
	\centering
	\setlength{\extrarowheight}{.5em}
	\scalebox{0.85}{
		\begin{tabular}{cccc}
			\toprule
			NUTS level  & Description & Number& Example (population)\\\hline
			\cellcolor{Gray}NUTS0 & \cellcolor{Gray}Country level & \cellcolor{Gray}1 & \cellcolor{Gray}AT Austria (8.86 million)\\
			NUTS1 & Major socioeconomic regions & 3 & AT3 Western Austria (2.78 million)\\
			NUTS2 & Basic regions for the application of regional policies (federal states) & 9 & AT31 Upper Austria (1.48 million)\\
			\cellcolor{Gray}NUTS3 & \cellcolor{Gray}(Small) sub-regions for specific diagnoses (political/court districts) & \cellcolor{Gray}35 & \cellcolor{Gray}AT312 Linz-Wels (529 thousand)\\
			\cellcolor{Gray}LAU (former NUTS4/5) & \cellcolor{Gray}Subdivision of the NUTS 3 regions (communities)& \cellcolor{Gray}2095 & \cellcolor{Gray}Enns AT312 Linz-Wels (11 thousand)\\ 
			\bottomrule
	\end{tabular}}
	\caption{Spatial nomenclature of different spatial levels using the NUTS nomenclature. Besides the number of regions per NUTS level, examples for the Austrian case study (incl. population) are given. The gray-colored rows mark the spatial levels used for downscaling in this work.}
	\label{tab:nuts}
\end{sidewaystable}

The purpose of the sequential downscaling algorithm is to provide a downscaling technique that considers the variation in efficiency of renewable heat sources\replaced{ and thus the prioritized use of heat sources/generation technologies in district heating.}{ and the increasing role of biomass and waste heat sources, in particular, in densely populated areas.} Hence, we claim that 

\begin{itemize}
	\item \added{geothermal sources significantly contribute to the decarbonization of heat demands but depend on the network infrastructure of district heating to be integrated into the heat supply \cite{weinand2019developing}}
	\item limited amounts of \added{synthetic gas and }hydrogen (\deleted{and/}or \replaced{\textit{green}}{"green"} gas) should preferably be used in district heating networks if they are used for heat supply \replaced{\cite{gerhardt2020hydrogen}}{(similar to Gerhardt et al.  \cite{gerhardt2020hydrogen} and Zwickl-Bernhard and Auer \cite{zwickl2021demystifying})}.
	\item high shares of biomass in the heating sector result in a high utilization rate of waste sources in waste incineration plants \cite{fruergaard2010energy}. These waste incineration plants feed into district heating and therefore depend on the infrastructure of \replaced{district}{centralized} heating networks \deleted{(see, e.g., Sahlin et al. }\cite{sahlin2004effects}\deleted{)}.
\end{itemize}

\deleted{Besides, we claim that high shares of air-source heat pumps (or geothermal sources) in the heat supply can only be realized if they are used as a co-firing heat source in district heating networks. We therefore consider two main aspects, namely that geothermal sources will contribute significantly to decarbonizing the feed-in energy mix of existing district heating grids in the future (see, e.g., \cite{weinand2019developing}), and that the provision of high shares of geothermal-based heat supply requires the distribution through district heating infrastructure \cite{dalla2020scenarios}. Besides, it is highly uncertain whether small-scale geothermal units at the end-user's level will be economically viable in the future, because of the high investment costs expected.}\vspace{0.3cm}

To incorporate the abovementioned relevant technology-specific aspects, heat technologies/sources are downscaled according to their necessity of distribution infrastructure. Therefore, population density serves as a criterion, indicating the possibility of \replaced{district heating}{centralized heat networks}. Table \ref{tab:require} provides a qualitative overview of the different heat sources/generation technologies and their heat network/infrastructure requirements.\vspace{0.3cm}

\newcommand{\xmark}{\ding{55}}
\definecolor{Gray}{gray}{0.95}
\begin{table}[h]
	\centering
	\setlength{\extrarowheight}{.5em}
	\resizebox{0.9\textwidth}{!}{
		\renewcommand{\arraystretch}{1.1}
		\begin{tabular}{lrcccl}
			\toprule
			\multicolumn{2}{c}{Heat supply}& \multicolumn{3}{c}{Type of sub-region (pop. density)} & \\\cmidrule(lr){3-5}\cmidrule(lr){0-1}
			\multirow{2}{*}{\makecell[l]{Source/\\technology}} & \multirow{2}{*}{\makecell[r]{Network\\requirement}} & \multirow{2}{*}{\makecell[c]{Rural\\(sparse)}} & \multirow{2}{*}{\makecell[c]{Town/Mixed\\(moderate)}} & \multirow{2}{*}{\makecell[c]{Urban\\(dense)}} & \multirow{2}{*}{\makecell[l]{Supporting\\references}}\\\\\hline
			Biomass & Low & \checkmark &  \checkmark & \checkmark& \\
			Direct electric & Low &  \checkmark & \checkmark & \checkmark&\\
			\cellcolor{Gray}Geothermal & \cellcolor{Gray}High &\cellcolor{Gray} & \cellcolor{Gray}& \cellcolor{Gray}\checkmark & \cellcolor{Gray}\cite{unternahrer2017spatial, weinand2019developing}\\
			Heat pump (air) & Low &  \checkmark & \checkmark & \checkmark&\\
			Heat pump (ground) & Low & \checkmark  &\checkmark   & \checkmark&\\
			\cellcolor{Gray}Hydrogen & \cellcolor{Gray}High & \cellcolor{Gray} &\cellcolor{Gray}  &  \cellcolor{Gray}\checkmark& \cellcolor{Gray}\cite{jensen2020potential, dodds2015hydrogen}\\
			\cellcolor{Gray}Synthetic gas & \cellcolor{Gray}High &\cellcolor{Gray} &\cellcolor{Gray} & \cellcolor{Gray}\checkmark& \cellcolor{Gray}\cite{jensen2020potential}\\
			\cellcolor{Gray}Waste & \cellcolor{Gray}High &\cellcolor{Gray} & \cellcolor{Gray}\checkmark & \cellcolor{Gray} \checkmark & \cellcolor{Gray}\cite{fruergaard2010energy, persson2016current}\\
			\bottomrule
	\end{tabular}}
	\caption{Qualitative overview of heat sources/generation technologies and their requirements for heat network infrastructure. The prioritized preferences of heat sources in sub-regions are marked by the gray cell. In addition, selected references supporting this assumptions are cited.}
	\label{tab:require}
\end{table}

\deleted{The sub-regions used to downscale the corresponding heat sources are marked. }Note that the different types are characterized by population density. Exemplarily, direct electric heating is a heat generation technology with no significant heat network requirements. It is downscaled to all types of sub-regions. In contrast, hydrogen is a heat source with high requirements and thus prioritized preferences (marked by the gray cell color). The right column refers to selected references whose key findings are in line with this approach/these assumptions. Building on this, the sequential downscaling algorithm is presented below (Algorithm 1).\vspace{0.6cm}

\scalebox{1}{
	\begin{algorithm}[H]
		\setstretch{1}
		\SetKwInOut{Input}{input}
		\SetKwInOut{Output}{output}
		\SetKwInput{kwInit}{Initialization}
		$t$: Heat generation by technology/source $(t \in \mathcal{T})$\;
		$r$: Sub-region (or NUTS3 region) $(r \in \mathcal{R})$\;
		\vspace{0.2cm}
		\Input{Heat generation by technology/source at NUTS0 level: $(q_{t})$;\newline
			Population density per sub-region $r$ $(\rho_{r})$;\newline
			Total population per sub-region $r$ $(p_{r})$;\newline
			Minimal network infrastructure requirements of $t$ $(\sigma_{t})$;\newline
			Available potential of heat network infrastructure at $r$ ($\pi_{r}$)\;}
		\vspace{0.2cm}
		\Output{Heat generation by technology/source at NUTS3 level $(\hat{q}_{t,r})$\;}
		\vspace{0.2cm}
		\kwInit{\\
			Sort elements $t$ in $T$ descending by $\sigma_{t}$\;
			$q^{heat}_{r} \longleftarrow \sum_{t} q_{t} \cdot \frac{p_{r}}{\sum_{r} p_{r}}$ \tcp*{Calculate heat demand at each sub-region}
			$\tilde{q}_{t} \longleftarrow q_{t}$ \tcp*{Available heat generation for each technology/source}
			$\pi_{r} \longleftarrow \rho_{r}$ \tcp*{Population density determines network potential}
			\vspace{0.2cm}}
		\SetAlgoLined
		\Begin{
			\ForEach{$t$}{
				$List=[~]$\tcp*{Collect valid sub-regions}
				$demand=0$\tcp*{Remaining demand that needs to be covered}
				$R^{'}=R\setminus \{\forall r \in R: \pi_{r} \leq \sigma_{t}\}$\tcp*{Get valid sub-regions by criteria}
				\ForEach{$r^{'} \in R^{'}$}{
					\If{$q^{heat}_{r} \ge 0$}{
						$List = List \cup r^{'}$\tcp*{Add valid sub-regions to collection} 
						$demand\mathrel{+}=q^{heat}_{r}$\tcp*{Total demand of valid sub-regions}}}
				\ForEach{$l \in List$}{
					$\hat{q}_{t,r} = \frac{q^{heat}_{r}}{demand}\cdot \tilde{q}_{t}$\tcp*{Population-based downscaling}
					$q^{heat}_{r} \mathrel{-}= \hat{q}_{t,r}$\tcp*{Reduce heat demand at $r$}
				}
			}
		}\caption{Sequential downscaling algorithm (NUTS0 to NUTS3)}
		\label{Alg:1}
\end{algorithm}}\vspace{0.6cm}

The inputs are as follows: (i) heat generation by technology/source at the NUTS0 level, (ii) population as well as population density at the NUTS3 level, and (iii) empirical assumptions in terms of network infrastructure requirements per heat technology/source and potentials for heat network infrastructure (see Table \ref{tab:require}). The algorithm itself consists of three main parts: initialization, pre-calculations, and downscaling. First, the initialization of the algorithm sorts the heat generation technologies/sources in descending order in terms of network infrastructure requirements. Then, the calculation starts with the first technology/source (highest requirements) (line 6). For this technology/source, all possible sub-regions are collected (line 9). Those sub-regions already fully supplied (no remaining heat demand) are filtered out (line 11). After further pre-calculation steps, the available amount of heat generation is downscaled to all valid sub-regions using population as a proxy. This procedure is repeated sequentially for each heat technology/source. The outputs of the sequential downscaling algorithm are heat generation by source and the amount of heat demand covered by centralized heat networks at the NUTS3 level.

\subsection{Iterative downscaling (from the sub-region to community levels)}\label{alg2}
This section explains the methodology of the iterative downscaling algorithm. We propose this downscaling technique projecting heat generation by technology/source from the sub-region (NUTS3) to the community levels (LAU) (see Table \ref{tab:nuts}). This in-depth spatial resolution is imperative for realistic network infrastructure planning\deleted{, as stated by Zvoleff et al.} \cite{zvoleff2009impact}. The underlying concept of iterative downscaling is based on graph theory and assessing network topology using benchmark indicators. 

\scalebox{1}{
	\begin{algorithm}[H]
		\setstretch{1}
		\SetKwInOut{Input}{input}
		\SetKwInOut{Output}{output}
		\SetKwInput{kwInit}{Initialization}
		$s$: Stage of iteration $(s \in \{0, 1, *\})$\;
		$G^{s}$: Centralized heat network graph at stage $s$\;
		$N^{s}$: List of nodes at stage $s$: ($n^{s} \in N^{s}$)\;
		$L^{s}$: List of lines connecting nodes $k$ and $j$ at stage $s$: ($l^{s}_{k,j} \in L^{s}$)\;
		$Q^{s}$: Centralized heat generation at stage $s$: $(q^{s}_{n^{s}} \in Q^{s})$\;
		$\tilde{Q^{s}}$: On-site heat generation at stage $s$: $(\tilde{q}^{s}_{n^{s}} \in \tilde{Q^{s}})$\;
		$\Pi^{s}$: Benchmark indicator value at stage $s$ ($\pi^{s}_{n^{s}} \in \Pi^{s}$)\;
		\vspace{0.2cm}
		\Input{$G^{0}=\{N^{0}, L^{0}, Q^{0}, \tilde{Q^{0}\}}$\;}
		\Output{$G^{*}=\{N^{*}, L^{*}, Q^{*}, \tilde{Q^{*}\}}$\;}
		\vspace{0.2cm}
		\kwInit{\\
			$s=0$, $iter=True$\;	
		}
		\SetAlgoLined
		\Begin{
			\While{$iter=True$}{
				\ForEach{$n \in N\textsuperscript{s}$}{
					$\Pi^{s}_{n^{s}}=f(N^{s}, L^{s}, Q^{s})$\tcp*{Calculate benchmark indicator value}}
				$i$ with $\pi^{s}_{i}=min(\Pi^{s}$)\tcp*{Get node with lowest indicator value}
				$N^{s+1}=N^{s} \setminus i$\tcp*{Remove node from graph obtaining next stage}
				$\tilde{q} = \sum_{N^{s+1}} \tilde{q}^{s}_{n^{s}}$\tcp*{Calculate available on-site heat generation}
				\eIf{$\tilde{q} \geq q^{s}_{i}$}{
					\ForEach{$n^{s+1}$}{
						$q^{s+1}_{n^{s+1}} = q^{s}_{n^{s}}+\frac{q^{s}_{i}}{\tilde{q}}\cdot \tilde{q}^{s}_{n^{s}}$\tcp*{Increase centralized heat amount}
						$\tilde{q}^{s+1}_{n^{s+1}} = \tilde{q}^{s}_{n^{s}}-\frac{q^{s}_{i}}{\tilde{q}} \cdot \tilde{q}^{s}_{n^{s}}$\tcp*{Decrease on-site heat amount}}
					$L^{s+1}=L^{s} \setminus \{\forall l^{s}_{k,j}: k=i \lor j=i\}$\tcp*{Remove connecting lines}
					$G^{s+1}=\{N^{s+1}, L^{s+1}, Q^{s+1}, \tilde{Q^{s+1}\}}$\tcp*{Create new network graph}
					$G^{s} = G^{s+1}$\tcp*{Set updated heat network graph as new input}
				}
			{
				$iterate=False$\tcp*{Stop iteration because of no reallocation}
				$G^{*} = G^{s}$\tcp*{Set heat network graph as result}}}}
		\caption{Iterative downscaling algorithm (NUTS3 to LAU level)}
		\label{Alg:2}
	\end{algorithm}
}\vspace{0.6cm}

\subsubsection{Algorithm description}
The iterative downscaling algorithm is presented in Algorithm 2. The idea is to assess, benchmark, and improve the topology of centralized heat networks. This is achieved in our proposed approach by iterative downscaling. Essentially, the main steps of the algorithm can be summarized as follows:

\begin{enumerate}[nolistsep]
	\item Downscale the results of the sequential downscaling algorithm from the NUT3 to the LAU levels using population as the downscaling driver, to obtain the initial heat network graph $G^{0}$ (input).
	\item Benchmark each node of the heat network graph (line 11), identify the node with the lowest indicator value, and remove the node from the graph, generating a reduced heat network graph (lines 13 and 14).
	\item Check if the amounts of centralized and on-site heat generation can be reallocated (line 16).
	\item If yes, reallocate centralized and on-site heat generation for all nodes (lines 18 and 19); otherwise stop algorithm.
	\item Update heat network graph and jump to step 2.
\end{enumerate}
\vspace{0.5cm}

Recent studies support this approach, focusing on the topography of energy systems and networks \deleted{(see, e.g., \cite{abuelnasr2018examining})}\added{\cite{abuelnasr2018examining}}. \replaced{Against this background, the study in \cite{bordin2016optimization} presents an optimization approach for district heating strategic network design. Further works also evaluate the impact of the heating system topology on energy savings \cite{allen2020evaluation}.}{Bordin et al. \cite{bordin2016optimization} conduct an approach for the optimized strategic network design of centralized heat systems. Allen et al. \cite{allen2020evaluation} evaluate the topology of centralized heating systems and conclude that the optimization of the topology is promising to facilitate the adoption of centralized heat networks.}

\subsubsection{Heat network topology benchmarking using graph theory}\label{bench}
So far, we have introduced only the function $f(N^{s}, L^{s}, Q^{s})$ (see line 11 in the iterative algorithm (Algorithm 2)) as a calculation procedure of the benchmarking indicator value. Below, we describe and discuss the approach of using a weighted cluster coefficient as a function and benchmarking indicator.\vspace{0.3cm}

The proposed benchmarking indicator value is derived from graph theory. Detailed information in the context of network analysis using indicators can be found in the fundamental work \replaced{in \cite{strogatz2001exploring}}{by Strogatz \cite{strogatz2001exploring}}. Morever, we refer the reader to \replaced{reference in}{Sanfeliu and Fu} \cite{sanfeliu1983distance}, which describes network topologies and their transformation in detail. In this work, we use a weighted cluster coefficient as a benchmark indicator and determine the transformation path of the centralized heat network graph. Equation \ref{eq:2} shows the calculation of the weighted cluster coefficient

\begin{align}\label{eq:2}
c_{n^{s}}=\frac{q_{n^{s}}}{max~q^{s}}\cdot \frac{\alpha_{n^{s}}}{\beta_{n^{s}}}
\end{align}

where $q$ is the amount of centralized heat supply, $\alpha$ is the number of triangles that can be formed with direct neighboring nodes, and $\beta$ is the number of lines connecting to the graph for node $n$ at stage $s$. In the context of the fundamental concept of $alpha$, we refer again to the literature. In particular, the study in \cite{huang2010link} comprehensively deals with cluster coefficients and provides related generalized concepts. In addition, relevant aspects of the cluster(ing) coefficient are shown in \cite{cui2014detecting}. In the works cited and also in this study, the aim is to achieve a high value of the cluster coefficient for each node considered (i.e., $\frac{\alpha}{\beta} \approx 1$). However, we extend the basic concept of the cluster coefficient from the literature and propose a weighting with the relative centrally supplied heat quantity. From an energy economics point-of-view, at least two important aspects are considered in the benchmarking process: (i) a high connection rate to the centralized heat network and (ii) a connection of those areas to the network that have a high heat demand and heat density, respectively. Both aspects are investigated in the literature \replaced{(connection rate in \cite{nilsson2008sparse} and the linearly decreasing heat densities and the influence on the profitability of district heating in \cite{dochev2018analysing}).}{For example, Nilsson et al. \cite{nilsson2008sparse} focus on the importance of the connection rate of centralized heat networks. Besides, Dochev et al. \cite{dochev2018analysing} investigate the impact of linearly decreasing heat densities and the influence on the profitability of the centralized heat networks.}

\subsection{\added{Limitation of the developed downscaling technique}}\label{sec:limitations}
\added{This section discusses selected limitations of the developed downscaling technique}\footnote{\added{Note that for the sake of convenience in this chapter we use the term technique and mean by this the toolkit of the sequential and the iterative downscaling technique together.}}. \added{In principle, various limitations of the developed downscaling technique could be identified. In the following section, we discuss three different aspects that are likely to be relevant in the context of heat supply in general, district heating and their heat densities.}

\subsubsection{Neglecting of heat sources on the local level due to the composition of the heat sector at the aggregated level}
\added{The developed downscaling technique disaggregates different heat sources/generation technologies to the local level based on their characterization in terms of heat network requirements and their prioritized use in district heating. Accordingly, it is advantageous to differentiate as many different heat sources/generation technologies as possible and take their special characteristics into account. This work essentially uses the different heat sources of the heating sector portfolio of Genesys-MOD which is why some heat sources are neglected. Exemplarily, industrial excess heat is not explicitly included in the values to be downscaled. And thus this heat source can not be reflected in the local heat supply and district heating. Adding information on the generation potential of industrial excess heat could improve the heat generation portfolio of heat sources to be downscaled, but also changes the inputs in terms of consistency of the downscaling and reduces the significance (bias by the new information) of the implications of the original input values.}

\subsubsection{\added{Downscaling global cost-effective heat generation values}}
\added{Although the values to be downscaled are the cost-effective solution of a decarbonized heat supply, the results of this work do not allow to assess cost-effective heat supply at the local level (i.e., cost-effective district heating networks). The main reason for this is that the benchmark indicator value calculation does not include infrastructure investment costs of district heating (incl. pipeline capacities based on the local peak heat demand). Note that the cost-effective solution is not in the focus of this work since the downscaling puts focus on the local implications of cost-effective national modeling results. Two more aspects are important here. Cost-effectivness of district heating is also influenced by the spatial distribution of heat sources which is neglected in the downscaling. And moreover that the spatial granularity is restricted to communities (LAU level). The estimates of heat densities of district heating could be improved by an even more higher spatial resolution within communities (i.e., street level).}

\subsubsection{\added{Influence of using population as a proxy on heat demands}}
\added{The developed downscaling technique uses population as a proxy for downscaling values from aggregated to local levels. With an eye on the heating sector, this could result in a bias in the estimates of local heat demands which have an impact on the obtained heat densities of district heating. Particularly, additional (low-temperature) heat demands in densely populated and urban areas are not considered. Moreover, the values to be downscaled are values per year which is why the peak heat demand is not considered in the analysis.}
	
\subsection{Development of an open-source package building on pyam}\label{open}
The method described will be released as an open-source Python package in the course of publishing this work at the author's GitHub account. In this package, we build on the existing open-source Python package \textit{pyam} \cite{huppmann2021pyam}. \textit{Pyam} is an open-source package for the analysis and visualization of integrated assessment and macro-energy scenarios. In this work, it is used particularly for (i) the linkage between the sequential and the iterative downscaling algorithms, (ii) the internal calculation steps within both downscaling algorithms, and (iii) the visualization of the results. Besides, we used the open-source Python package \textit{networkx} \cite{hagberg2008exploring}, when implementing the iterative downscaling algorithm. We refer to the repository for the codebase, data collection, and further information. 
