\section{Conclusions and outlook}\label{conclusions}
The sustainable energy transition requires methods to bridge the gap between global decarbonization plans and the resulting necessary measures at the local level. This work emphasizes downscaling of the Austrian heating sector under the 1.5°C climate target to the grid level, considering technology-specific infrastructure requirements for the highly efficient usage of heat sources.\vspace{0.3cm}

We found that the prioritized perspective of efficiency and local utilization of renewable heat sources leads to a crucial role of district heating networks in the decarbonized Austrian heat supply 2050. It is shown that this implies small-scale ($<\SI{1}{TWh}$) and large-scale ($>\SI{6}{TWh}$) district heating networks in terms of the amount of heat delivered. The results demonstrate that particularly densely populated areas are beneficial supply areas for district heating networks and offer adequate heat densities. Nevertheless, most district heating networks in 2050 (five out of six) will not reach the heat densities of today's networks and have a significant heat density gap. We identify heat densities there of \SI{2.58}{\frac{GWh}{km^2}} and less.\vspace{0.3cm}

We anticipate our work as a starting point discussing the role of centralized heat networks in the light of enabling large-scale, highly efficient, and local integration of renewable heat sources (such as biomass/waste and hydrogen). In particular, we see a need for further research on the trade-off analysis between the efficiency/local integration of heat sources and the cost-intensive deployment of district heating networks.  








%The sustainable energy transition requires methods to bridge the gap between global plans and implications and processes on the local level. Techniques for downscaling of global decarbonization scenarios to finer scales will become increasingly important in the future. Thereby, energy-policy makers have to rely on the meaningfulness of the downscaled values, which requires tailor-made downscaling techniques for the different fields of energy systems. This work emphasizes a downscaling technique for the residential and commercial heating sector, taking into account the technology-specific requirements of heat network infrastructure accounting for the highly efficient use of energy carriers. In particular, the proposed downscaling techniques reveal the potentials of centralized heat supply in four different European decarbonization scenarios on a high spatial granularity.\newline
%
%Results indicate that centralized heat systems undergo a fundamental shift going beyond the decarbonization of the supplying energy mix. In particular, the reduction of heat density of centralized heat networks compared to today's networks poses massive challenges to heat supply companies and fundamentally jeopardizes associated business models. At the same time, however, the heat network infrastructure may play a crucial role in expected energy systems since both use of local energy sources efficiently and on a large scale (e.g., geothermal sources, waste incineration, waste heat from industry, etc.) and unburden the electricity sector taking into account the aim of high electrification of different energy services. These trade-offs should be given greater consideration in the future and may have implications for the regulation and benchmarking of heat supply companies that provide centralized heat network infrastructure.\newline
%
%Future work may address improvements of the proposed downscaling technique for the heating sector, taking into account a finer scale of spatial granularity, extensions of the introduced benchmarking of centralized heat networks using indicator values in the context of heat sources and local characteristics, higher resolution of the heat generation technologies/sources in terms of requirements for heat network infrastructure, and a detailed cost-benefit analysis of the centralized heat systems obtained by the downscaling (e.g., distribution line capacities, connection capacities to the public grid, etc.). 


%We anticipate our work as a starting point for more sophisticated models 
%"Future work" einbetten.

