\documentclass[review]{elsarticle}
\input{packages.tex}

\begin{document}
\begin{frontmatter}

\title{Downscaling European decarbonization scenarios of the heating sector to the Austrian community level: Assessing the heat density gap of centralized heat networks between 2050 and today}
\author[1,2]{Sebastian Zwickl-Bernhard\corref{cor1}}
\ead{zwickl@eeg.tuwien.ac.at}
\author[2]{Daniel Huppmann}
\author[1]{Antonia Golab}
\author[1]{Hans Auer}
\cortext[cor1]{Corresponding author}
\address[1]{Energy Economics Group (EEG), Technische Universität Wien, Gusshausstrasse 25-29/E370-3, 1040 Wien, Austria}
\address[2]{Energy, Climate and Environment (ECE) Program,  International Institute for Applied Systems Analysis (IIASA), Laxenburg, Austria}

\begin{abstract}
	The core objective of this work is to downscale European deep decarbonization scenarios of the heating sector to the Austrian community level in 2050 to obtain an estimation of future demand for different heating technologies on smaller geographical scale. We use tailor-made downscaling techniques accounting for infrastructure needs of heat sources, the network topology of centralized heat networks, and population density. The results show that centralized heat networks play an important role in the highly efficient heat supply of densely populated areas in 2050. However, the projected heat density of supply areas is significantly lower than today despite high connection rates (in average greater than \SI{85}{\%}). We identify a heat density gap of at least \SI{7.42}{\frac{GWh}{km^2}} considering \SI{10}{\frac{GWh}{km^2}} as today's minimum required value. We conclude that future yes/no decisions of centralized heat networks should take enhanced account of the highly efficient use of sustainable and local heat sources in addition to the achievable heat density. Policy-makers should incorporate the development of lower heat densities resulting from decarbonization measures in the benchmarking of centralized heat networks efficiency.  
\end{abstract}

\begin{keyword}
	Downscaling, Centralized heat networks, Heat density, Network topology, Infrastructure requirements
\end{keyword}
\end{frontmatter}

\linenumbers
\input{1_2_Introduction.tex}
\newpage
\input{3_Methodology.tex}
\newpage
\input{4_Results.tex}
\newpage
\input{5_Conclusions.tex}




\section*{Declaration of interests}
None.
\section*{Declaration of Competing Interest}
The authors report no declarations of interest.
\section*{Acknowledgments}
This project has received funding from the European Union's Horizon 2020 Research and Innovation Programme under Grant Agreement No. 835896. Part of the research was developed in the
Young Scientists Summer Program (YSSP) at the International Institute for Applied Systems Analysis(IIASA), Laxenburg (Austria).

\bibliography{mybibfile}
\appendix
\setcounter{table}{0}
\setcounter{figure}{0}


\section{Data and further empirical settings}\label{appendixA}
\begin{table}[h]
	\centering
	\scalebox{0.85}{
		\renewcommand{\arraystretch}{1.35}
		\begin{tabular}{lccc}
			\toprule 
			& Description & Data availability & Data source\\\hline
			GENeSYS-MOD v2.0 & Heat generation by source & \cite{explorer} & \cite{loffler2017designing}\\
			Austrian population density & in 2019 & \href{https://www.statistik.at/web_de/statistiken/index.html}{\textit{Statistik Austria}} & as availability\\
			Austrian population & in 2050 & \href{https://ec.europa.eu/eurostat/databrowser/view/tps00003/default/table?lang=en}{\textit{Eurostat}} & as availability\\
			\bottomrule
	\end{tabular}}
	\caption{Empirical data settings}
	\label{tab:a2}
\end{table}


\end{document}